---
title: "SIOPE API – Education spending data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIOPE API – Education spending data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## About

SIOPE (Sistema de Informações sobre Orçamentos Públicos em Educação)
provides education spending data from states and municipalities. The
system is maintained by FNDE (Fundo Nacional de Desenvolvimento da
Educação) and the API uses an OData-style interface hosted at
`https://www.fnde.gov.br/olinda-ide/servico/DADOS_ABERTOS_SIOPE/`.

Data covers revenues, expenses, indicators, official contacts, and
staff compensation related to public education spending.

## Available functions

| Portuguese | English | Description |
|:---|:---|:---|
| `get_siope_dados_gerais()` | `get_siope_general_data()` | General data (demographics, GDP, totals) |
| `get_siope_responsaveis()` | `get_siope_officials()` | Officials responsible for declarations |
| `get_siope_despesas()` | `get_siope_expenses()` | Education expenses |
| `get_siope_despesas_funcao()` | `get_siope_expenses_by_function()` | Expenses by government function |
| `get_siope_indicadores()` | `get_siope_indicators()` | Education spending indicators |
| `get_siope_info_complementares()` | `get_siope_supplementary()` | Supplementary information |
| `get_siope_receitas()` | `get_siope_revenues()` | Education revenues |
| `get_siope_remuneracao()` | `get_siope_compensation()` | Staff compensation |

## Required parameters

All functions share three required parameters:

| Portuguese | English | Description | Example |
|:---|:---|:---|:---|
| `ano` | `year` | Year of the data | `2023` |
| `periodo` | `period` | Bimester (1-6) | `6` (Nov-Dec) |
| `uf` | `state` | State abbreviation | `"PE"` |

**Exception**: `get_siope_remuneracao()` / `get_siope_compensation()`
has a 4th required parameter: `mes` / `month` (1-12).

### Period mapping

| Period | Months |
|:---|:---|
| 1 | January – February |
| 2 | March – April |
| 3 | May – June |
| 4 | July – August |
| 5 | September – October |
| 6 | November – December |

## Performance tip: use `filter` to reduce downloads

> **The SIOPE API returns all municipalities for a given state.**
> A single call for São Paulo (`state = "SP"`) returns 645+ rows.
> Use the `filter` parameter to apply **server-side filtering** and
> download only the data you need. This is much faster than downloading
> everything and filtering locally with `dplyr::filter()`.

### Filtering by municipality

```{r}
# SLOW: downloads all 185 municipalities, then filters locally
dados_pe <- get_siope_general_data(year = 2023, period = 6, state = "PE")
recife <- dplyr::filter(dados_pe, nom_muni == "Recife")

# FAST: server returns only Recife's data
recife <- get_siope_general_data(
  year = 2023, period = 6, state = "PE",
  filter = "NOM_MUNI eq 'Recife'"
)

# Filter by IBGE code
recife <- get_siope_general_data(
  year = 2023, period = 6, state = "PE",
  filter = "COD_MUNI eq 261160"
)
```

### OData filter syntax

The `filter` parameter uses OData v4 syntax. Column names must use the
**original API names** (uppercase), not the cleaned snake_case names.
Use `verbose = TRUE` on a `max_rows = 1` call to inspect column names.

| Operator | Meaning | Example |
|:---|:---|:---|
| `eq` | Equals | `"NOM_MUNI eq 'Recife'"` |
| `ne` | Not equals | `"TIPO ne 'Estado'"` |
| `gt`, `ge` | Greater (or equal) | `"NUM_POPU gt 100000"` |
| `lt`, `le` | Less (or equal) | `"NUM_POPU le 50000"` |
| `and` | Logical AND | `"NOM_MUNI eq 'Recife' and TIPO eq 'Município'"` |
| `or` | Logical OR | `"NOM_MUNI eq 'Recife' or NOM_MUNI eq 'Olinda'"` |
| `contains()` | Substring match | `"contains(NOM_MUNI, 'Porto')"` |

### Selecting specific columns

Use `select` to request only the columns you need (reduces payload):

```{r}
# Only municipality name and declared value
resumo <- get_siope_expenses(
  year = 2023, period = 6, state = "PE",
  select = c("NOM_MUNI", "NOM_ITEM", "VAL_DECL"),
  filter = "NOM_MUNI eq 'Recife'"
)
```

> **Column names in filter/select/orderby must use the original API
> names (UPPERCASE), not the cleaned snake_case names.** To discover
> valid column names, run a quick test:
>
> ```r
> sample <- get_siope_expenses(
>   year = 2023, period = 6, state = "PE", max_rows = 1
> )
> # snake_case names → convert back to UPPER for filter/select
> toupper(names(sample))
> ```

### Sorting results

```{r}
# Sort by population descending
dados <- get_siope_general_data(
  year = 2023, period = 6, state = "PE",
  orderby = "NUM_POPU desc", max_rows = 10
)
```

### Compensation data with filters

```{r}
# Staff compensation for Agrestina, only "Efetivo" professionals
rem <- get_siope_compensation(
  year = 2024, period = 1, month = 1, state = "PE",
  filter = "NOM_MUNI eq 'Agrestina' and DS_SITUACAO_PROFISSIONAL eq 'Efetivo'"
)
```

## Examples

```{r}
library(tesouror)
library(dplyr)

# General data for all municipalities in Pernambuco, last bimester 2023
dados_pe <- get_siope_general_data(year = 2023, period = 6, state = "PE")

# Education revenues for São Paulo
receitas_sp <- get_siope_revenues(year = 2023, period = 6, state = "SP")

# Education indicators for Minas Gerais
indicadores_mg <- get_siope_indicators(year = 2023, period = 6, state = "MG")

# Expenses by function for Rio de Janeiro
desp_func_rj <- get_siope_expenses_by_function(
  year = 2023, period = 6, state = "RJ"
)

# Staff compensation for December 2023 in Bahia
remuneracao_ba <- get_siope_compensation(
  year = 2023, period = 6, month = 12, state = "BA"
)
```

### Quick test with max_rows

```{r}
# Grab just 10 rows to inspect the structure
sample <- get_siope_general_data(
  year = 2023, period = 6, state = "PE", max_rows = 10
)
glimpse(sample)
```

### Combining states

```{r}
# Fetch multiple states and combine
nordeste <- c("AL", "BA", "CE", "MA", "PB", "PE", "PI", "RN", "SE")

indicadores_ne <- purrr::map_dfr(nordeste, function(uf) {
  get_siope_indicators(year = 2023, period = 6, state = uf)
})
```

## How the OData URL works

The SIOPE API uses an OData-style URL pattern different from the other
APIs in this package. Internally, the package builds URLs like:

```
{base}/Dados_Gerais_Siope(Ano_Consulta=@Ano_Consulta,Num_Peri=@Num_Peri,Sig_UF=@Sig_UF)
  ?@Ano_Consulta=2023&@Num_Peri=6&@Sig_UF='PE'&$format=json
```

where `{base}` is the FNDE/MEC SIOPE OData root, available from
`tesouror:::siope_base_url()`.

You don't need to worry about this — just pass the parameters and the
package handles the rest. Use `verbose = TRUE` to see the full URL:

```{r}
get_siope_general_data(year = 2023, period = 6, state = "PE", verbose = TRUE)
#> ℹ API call: https://www.fnde.gov.br/olinda-ide/servico/...
```
